\GVN{}-Hoist: Hoisting Computations from Branches

### 2016-08-25 hoist.tex
more cleanups

\begin{document}
\def \SCoP {SCoP}
\def \GCC {GCC}
\def \LLVM {LLVM}
\def \SESE {SESE}
\def \CFG {CFG}
\def \SSA {SSA}
\def \scev {scev}

@@

\title{GVN-Hoist: Hoisting Computations from Branches}

\toappear{
   \hrule \vspace{5pt}
   LLVM-dev 2016
}
\numberofauthors{2}

@@

  We present a code hoisting pass that we implemented in LLVM. It is based on
  Global Value Numbering infrastructure available in LLVM. The experimental
  results show an average of 2.5\% savings in code size, although the code size
  increases in many cases because it enables more inlining. This is an
  optimistic algorithm in the sense that we consider all identical computations
  in a function as potential candidates to be hoisted. We make an extra effort
  to hoist candidates by partitioning the potential candidates in a way to
  enable partial hoisting in case common hoisting points for all the candidates
  cannot be found. We also formalize cases when register pressure will reduce as
  a result of hoisting.
\end{abstract}

@@

Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like CSE \cite{dragonbook} work very well on single basic blocks.
When it comes to detect redundancies across basic blocks these techniques fall
short: more complex passes like GCSE and PRE have been designed to handle these
cases based on dataflow analysis \cite{morel1979global}.  At first these
techniques were described in the classical data-flow analysis framework, and
later the use of the SSA representation lowered the cost in terms of compilation
time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
techniques in the main stream: nowadays SSA based PRE is available in every
industrial compiler.

This paper describes code-hoisting, a technique that uses the information
computed for PRE to detect identical computations and has a transformation phase
whose goal differs from PRE: it removes identical computations from different
branches of execution.  These identical computations in different branches of
execution are not redundant computations at run-time and the number of run-time
computations is not reduced. Code-hoisting is not a redundancy elimination pass,
and thus it has different cost function and heuristics than PRE or CSE.
Code-hoisting is also different from global code scheduling
\cite{dragonbook,click1995global} in the sense that code-hoisting will only
hoist computations when there are identical computations sharing a common
dominator. The goals of code hoisting are:

@@

The main contributions of this paper are:
\begin{itemize}
\item a new algorithm to hoist computations from branches,
\item cost models to reduce live-range and reduce spills,
\item performance evaluation of the implementation in LLVM,
\item comparison against other algorithms for code hoisting.
\end{itemize}

@@

There are a lot of bug reports in GCC and LLVM bugzillas
\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current LLVM implementation of code
hoisting in SimplifyCFG.cpp $HoistThenElseCodeToIf()$ is very limited to
hoisting from identical basic blocks: the instructions of two sibling basic
blocks are read in the same time, and all the instructions of the blocks are
hoisted to the common parent block as long as the compiler is able to prove that
the instructions are equivalent.  This implementation does not allow for an easy
extension: first in terms of compilation time overhead the implementation would
become quadratic in number of instructions to bisimulate, and second the
equivalence of instructions currently uses $$I1->isIdenticalToWhenDefined(I2)$$
would need to be rewritten to be more general, leading to using a mechanism
similar to the idea described in this paper based on GVN.

@@

Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
are computed which are hoisted in a basic block where the expression is
anticipable (all the operands are available). This algorithm would hoist as far
as possible without regarding the impact on register pressure and as such a cost
model will be required.  Also the description of VBE is based on the classic
dataflow model and an adaptation to a sparse SSA representation is required.

GCC recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
part of GVN-PRE: it uses the set of ANTIC and AVAIL value expressions computed
for PRE.  The algorithm hoists top down in a predecessor where the value is in
ANTIC.  It uses ANTIC[B] to know what expressions will be computed on every
path from B to exit, and can be computed in B: that is the safety condition.  It
uses AVAIL[B] to subtract out those values already being computed (because they
are already available): this is a redundancy condition.  The cost function is:
for each hoist candidate, if all successors of B are dominated by B, then we
know insertion into B will eliminate all the remaining computations.  It then
checks to see if at least one successor of B has the value available.  This
avoids hoisting it way up the chain to ANTIC.  It also checks to ensure that B
has multiple successors, since hoisting in a straight line is pointless.  The
algorithm continues on down the dominator tree, iterating with PRE until no more
changes.  One advantage of GCC implementation is that it works in sync with the
GVN-PRE such that when new hoisting opportunities are created by GVN-PRE,
code-hoisting will hoist them.

@@

The algorithm for code hoisting uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Control Flow Graph (CFG) and the Dominance (DOM) and Post-Dominance (PDOM)
  relations \cite{dragonbook};
\item Single Entry Single Exit (SESE) \cite{sese} and Single Entry Multiple Exit
  (SEME) regions;
\item Static Single Assignment (SSA) \cite{cytron};
\item Global Value Numbering (GVN) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use GVN.  Each expression is given a
  unique number and the expressions that the compiler can prove identical are
  given the same number;
\item Memory SSA \cite{novillo2007memory}: memory operations that the compiler
  is able to prove in dependence are linked through use-def chains.
\end{itemize}

@@

\subsection{Finding candidates to hoist}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by GVN.

The current implementation of GVN in LLVM has some limitations when it comes to
loads and stores so we compute the GVN of loads and stores separately.  Our
short-time solution to value number loads is to hash the address from where the
value is to be loaded. For stores, we value number the address as well as the
value to be stored at that address.

Another limitation of the current GVN implementation in LLVM is that the
instructions dependent on the loads will not get numbered correctly, and so
after hoisting all candidates we need to rerun the GVN analysis in order to
discover new candidates now available after having hoisted load instructions.
This limitation should be addressed in a new implementation of the GVN based on
MemorySSA, that would better account for equivalent loads and their dependent
instructions.

The process of computing GVN can be on-demand (as we come across an instruction)
or, precomputed (computing GVN of all the instructions beforehand). Which
process to choose is determined by the scope of code-hoisting we want to
perform. In a pessimistic approach, as described in
Section~\ref{subsec:pessimistic}, we want to hoist a limited set of instructions
from the sibling branches as we iterate the DFS tree bottom-up, it is sufficient
to compute GVN values on-demand. Whereas, in the optimistic approach, as
described in Section~\ref{subsec:optimistic}, we want to hoist as many
instructions as possible, and it would require GVN values to be precomputed.

@@

\subsubsection{Profitability of hoisting scalars}
Since scalars are the majority of instructions which are hoisted, we pay special
attention in case of hoisting scalars too far, as that may increase register
pressure and result in spills. For example hoisting a scalar past a call, as
described in Section~\ref{cost:across-calls}.  In our current implementation we
hoist scalars past a call only when optimizing for code-side (-Os). Ideally, a
later stage of live-range splitting pass should split the live-ranges for
optimal performance, however, that is not the case with llvm as we have found
regressions when scalars are hoisted too far, as in
Section~\ref{sec:cost-models}. Another way to mitigate this problem is be to
reinstantiate (rematerialize) the computation after a call (may be as a
different optimization pass).

@@

\subsection{Code generation}
Once all the legality and profitability checks are satisfied for a set of
identical instructions, they are suitable candidates for hoisting. A copy of the
computation is inserted at the hoisting point along with any instructions which
needed to be rematerialized. Thereafter, all the computations made redundant by
the new copy are removed, and the SSA form is restored by updating the
intermediate representation (IR) to reflect the changes.

After one iteration of algorithm runs through the entire function, it creates
more oppportunities for \emph{higher ranked} computations
\cite{rosen1988global}. Currently, this is a limitation of the GVN analysis
pass, and so we rerun the code-hoisting algorithm until there are no more
instructions left to be hoisted.  Obviously, this is not the most optimal
approach and can be improved by ranking the computations \cite{rosen1988global},
or by improving the GVN analysis to correctly number loads and dependent
instructions.

Finally after the transform is done, we verify a set of post-conditions to
establish that program invariants are maintained: e.g., consistency of
use-defs, and SSA semantics.

@@

In this program the computations of tmax and tmin are identical to the
computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
depends on inv which depends on a division operation which is generally more
expensive than the addition, subtraction and multiplication operations. The
total latency of computation across each branch is:
$C(div) + 2(C(sub) + C(mul))$
Or, for out of order processors with two add units and two multiply units:
$C(div) + C(sub) + C(mul)$

@@

In this code the two subtractions and the division operations can be executed in
parallel because there are no dependencies among them. So the total number of
cycles will be $max(C(div), C(sub)) + C(mul) = C(div) + C(mul)$; since $C(div)$ is
usually much greater than $C(sub)$ \cite{x86,aarch64}.

A partial redundancy elimination pass could now remove the entire if-block
because final operation is an addition (asociative under fast math), and the end
result is:

\begin{verbatim}
float fun(float d, float min, float max, float a)
{
  float tmin1, tmax1, inv;

  return tmax1 + tmin1;
}
\end{verbatim}

@@

\section{Code hoisting policies}
The amount of hoisting depends on whether we collect GVN of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the GVN algorithm, however, that analysis is beyond
the scope of this paper.

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
In this approach, the goal is to maximize the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We collect the GVN of all the instructions in the function and iterate on the
list of instructions having identical GVNs. After that we find the common
dominator dominating all such identical computations and perform legality
checks, as described in Section~\ref{subsec:legality}. Often times it is not
possible to hoist all the instructions to one common dominator, due to legality
constraints e.g., intersecting side-effects or, profitability constraints e.g.,
hoisting point too far. In those cases, this algorithmn would partition the list
of identical instructions into subsets which can be partially hoisted to their
respective common dominators. The partition algorithm is described as follows:

@@

Same property would hold for instructions I1, I2, I3 provided they are not fully
redundant i.e., none of the basic blocks containing I1, I2 and I3 respectively
dominate the other:

@@

In the presence of fully redundant computations in the list of hoistable
candidates this equation may not hold. For example, if there is another
instruction I4 which is dominated by I1, but I4 is present in a basic block
which is farther than DFSIn(I1, I2) i.e., $DFSIn(I1,I2) < DFSIn(I1,I4)$;
however, $BBDist(I1,I4) = 0 < BBDist(I1,I2) >=1$.

@@

In our current implementation we keep as many candidates in one set as possible
(greedy approach). We split the list at a point where the legality checks fail
to hoist subset of candidates which are legal to hoist and then start finding
new hoisting point for the remaining ones. So there are two limitations of the
current implementation of the partition algorithm: First, sorting by DFSIn
numbers does not give the desired order when there are fully redundant
instructions; Second, it lacks cost model for partitioning in order to
preserve/improve performance, see Section~\ref{sec:future-work}.

\subsection{Hoisting bottom up with pessimistic approach}
\label{subsec:pessimistic}
In the pessimistic approach, the basic blocks are traversed in the inverse
depth-first order, computing the GVN of instructions as they come by. The GVN of
sibling branches are compared for equality. Once such a candidate is found, it
is hoisted in the common dominator. All the leaf nodes are visited before the
non-leaf nodes (bottom-up) because instructions are hoisted upwards.

The pessimistic algorithm is fast, results in fewer spills but hoists very less
instructions. We have implemented optimistic approach because it is more general
and can be tuned down to closely mimic pessimistic approach by changing a flag.

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that could
be hoisted in the program, matching the complexity of PRE on SSA form.  The
analysis phase is based on the Global Value Numbering (GVN), the same analysis
used for PRE, followed by the computation of a partition of identical
expressions to be hoisted in a same location to guarantee safety properties and
program performance, and followed by a simple code generation that adds the
identified instruction in the hoisting point and removes all the now redundant
expressions.

@@

If the new live-range is less than the old one it will be a good candidate for
hoisting. The live ranges can remain same as well if there is only one operand
on the right hand side e.g., an assignment operation or, one of the operands is
not a register. That means hoisting upwards will decrease the live-range of its
use but increase the live-range of its definition.

@@

\subsection{In the presence of calls}
\label{cost:across-calls}
Hoisting scalars across calls is tricky because it can increase the number of
spills. During the frame lowering of calls, the argument registers, in general,
the caller saved registers are saved because they might be modified by the
callee and after the call they are restored \cite{frame-lowering}. So before the
call, the register pressure is high because the number of available registers
are reduced by the number of caller saved registers. In that situation if a
computation is hoisted across the call, that would increase the total number of
registers required by one, thus contributing to the register pressure. However,
in the special case discussed in Section~\ref{hoist:reg-pressure}, it will be
okay to hoist because the register pressure would not decrease.

Hoisting loads/stores across calls also require precise analysis of all the
memory addresses accessed by the call. Our implementation being an
intraprocedural pass, the analysis is very conservative. In the presence of pure
calls, loads can be hoisted but stores can't. Also, if the call throws
exceptions, or it it may not return, memory references cannot be hoisted.

\subsection{Hoisting too far away}
If there are several instructions in between the hoisting point and the
instruction to be hoisted, the instruction to be hoisted crosses several
instructions while hoisting, that means we are adding one register to all the
live-ranges spanning the instructions. That could result in spills. In the
current implementation we choose to hoist if the number of instructions crossed
is below a threshold. Ideally, it should be okay to hoist all the instructions
and a later a live-range-splitting \cite{cooper1998live} pass should make the
right decision of rematerializing the instruction should it be beneficial to do
so. But the current live-range splitting pass of llvm is not making the optimial
decision and we have found spills if the threshold is exceeded. The threshold
was computed as a result of tuning the llvm testsuite \cite{llvm-nightly} and
spec benchmark \cite{Henning2000}.

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
We ran llvm-testsuite (trunk:d87471f8) with the patch (trunk:86940146) and the
results are listed in Table~\ref{tab:hoist-results}. The table lists the number
of scalars, loads, stores and calls hoisted as well as removed. For each
category, the number of instructions removed is greater or equal to the number
of instructions hoisted because each hoisting is performed only when at least
one identical computation is found.

Loads are hoisted the most followed by scalars, stores and calls in decreasing
order.  This was the common trend in all our experiments. One reason why loads
are hoisted the most is the early execution of this pass (before mem2reg) in the
llvm pass pipeline. Passes like mem2reg, instcombine might actually remove
those loads so this order may change should this pass be scheduled later.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|}
      \hline
      Metric               & Number\\\hline
      Scalars hoisted      & 6791  \\\hline
      Scalars removed      & 9696  \\\hline
      Loads hoisted        & 14802 \\\hline
      Loads removed        & 20719 \\\hline
      Stores hoisted       & 15    \\\hline
      Stores removed       & 15    \\\hline
      Calls hoisted        & 8     \\\hline
      Calls removed        & 8     \\\hline
      Total Instructions hoisted & 21616 \\\hline
      Total Instructions removed & 30438 \\\hline
\end{tabular}
  \end{center}
  \caption{Code hoisting metrics on llvm-testsuite}
  \label{tab:hoist-results}
\end{table}

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|}
      \hline
      Metric               & Before & After              \\\hline
      Call sites deleted, not inlined             & 1988    & 1988   \\\hline
      Functions deleted (all callers found)       & 38250   & 38255  \\\hline
      Functions inlined                           & 154986  & 154985 \\\hline
      Allocas merged together                     & 212     & 212    \\\hline
      Caller-callers analyzed                     & 193042  & 193092 \\\hline
      Call sites analyzed                         & 414336  & 414381 \\\hline
      Rematerialized defs for spilling            & 18321   & 18326  \\\hline
      Rematerialized defs for splitting           & 5719    & 5842   \\\hline
      Spill slots allocated                       & 42912   & 42970  \\\hline
      Spilled live ranges                         & 61330   & 61362  \\\hline
      Spills inserted                             & 50724   & 50784  \\\hline
\end{tabular}
  \end{center}
  \caption{Static metrics before and after code-hoisting on llvm-testsuite}
  \label{tab:static-results}
\end{table}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|}
      \hline
      Code-size metric  (.text)                   & Number   \\\hline
      Total  benchmarks                           & 497      \\\hline
      Total  gained in size                       & 39       \\\hline
      Total  decrease in size                     & 58       \\\hline
      Median decrease in size                     & 2.9\%    \\\hline
      Median increase in size                     & 2.4\%    \\\hline
    \end{tabular}
  \end{center}
  \caption{Code size metrics on llvm-testsuite}
  \label{tab:code-size}
\end{table}

>>>>

\begin{document}
\def \GCC {GCC}
\def \LLVM {LLVM}
\def \SESE {SESE}
\def \CFG {CFG}
\def \SSA {SSA}
\def \MemorySSA {MemorySSA}
\def \PRE {PRE}
\def \GVN {GVN}
\def \LLVMTestSuite {\LLVM{} test-suite}

@@

\title{\GVN{}-Hoist: Hoisting Computations from Branches}

\toappear{
   \hrule \vspace{5pt}
   \LLVM{}-dev 2016
}
\numberofauthors{2}

@@

  We present a code hoisting pass that we implemented in \LLVM{}. It is based on
  Global Value Numbering infrastructure available in \LLVM{}. The experimental
  results show an average of 2.5\% savings in code size, although the code size
  increases in many cases because it enables more inlining. This is an
  optimistic algorithm in the sense that we consider all identical computations
  in a function as potential candidates to be hoisted. We make an extra effort
  to hoist candidates by partitioning the potential candidates in a way to
  enable partial hoisting in case common hoisting points for all the candidates
  cannot be found. We also formalize cases when register pressure will reduce as
  a result of hoisting.
\end{abstract}

@@

Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like CSE \cite{dragonbook} work very well on single basic blocks.
When it comes to detect redundancies across basic blocks these techniques fall
short: more complex passes like GCSE and \PRE{} have been designed to handle these
cases based on dataflow analysis \cite{morel1979global}.  At first these
techniques were described in the classical data-flow analysis framework, and
later the use of the \SSA{} representation lowered the cost in terms of compilation
time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
techniques in the main stream: nowadays \SSA{} based \PRE{} is available in every
industrial compiler.

This paper describes code-hoisting, a technique that uses the information
computed for \PRE{} to detect identical computations and has a transformation phase
whose goal differs from \PRE{}: it removes identical computations from different
branches of execution.  These identical computations in different branches of
execution are not redundant computations at run-time and the number of run-time
computations is not reduced. Code-hoisting is not a redundancy elimination pass,
and thus it has different cost function and heuristics than \PRE{} or CSE.
Code-hoisting is also different from global code scheduling
\cite{dragonbook,click1995global} in the sense that code-hoisting will only
hoist computations when there are identical computations sharing a common
dominator. The goals of code hoisting are:

@@

The main contributions of this paper are:
\begin{itemize}
\item a new algorithm to hoist computations from branches,
\item cost models to reduce live-range and reduce spills,
\item performance evaluation of the implementation in \LLVM{},
\item comparison against other algorithms for code hoisting.
\end{itemize}

@@

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of code
hoisting in SimplifyCFG.cpp $HoistThenElseCodeToIf()$ is very limited to
hoisting from identical basic blocks: the instructions of two sibling basic
blocks are read in the same time, and all the instructions of the blocks are
hoisted to the common parent block as long as the compiler is able to prove that
the instructions are equivalent.  This implementation does not allow for an easy
extension: first in terms of compilation time overhead the implementation would
become quadratic in number of instructions to bisimulate, and second the
equivalence of instructions currently uses $$I1->isIdenticalToWhenDefined(I2)$$
would need to be rewritten to be more general, leading to using a mechanism
similar to the idea described in this paper based on \GVN{}.

@@

Dhamdhere \cite{dhamdhere1988fast}, Muchnick \cite{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
are computed which are hoisted in a basic block where the expression is
anticipable (all the operands are available). This algorithm would hoist as far
as possible without regarding the impact on register pressure and as such a cost
model will be required.  Also the description of VBE is based on the classic
dataflow model and an adaptation to a sparse \SSA{} representation is required.

\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
part of \GVN{}-\PRE{}: it uses the set of ANTIC and AVAIL value expressions computed
for \PRE{}.  The algorithm hoists top down in a predecessor where the value is in
ANTIC.  It uses ANTIC[B] to know what expressions will be computed on every
path from B to exit, and can be computed in B: that is the safety condition.  It
uses AVAIL[B] to subtract out those values already being computed (because they
are already available): this is a redundancy condition.  The cost function is:
for each hoist candidate, if all successors of B are dominated by B, then we
know insertion into B will eliminate all the remaining computations.  It then
checks to see if at least one successor of B has the value available.  This
avoids hoisting it way up the chain to ANTIC.  It also checks to ensure that B
has multiple successors, since hoisting in a straight line is pointless.  The
algorithm continues on down the dominator tree, iterating with \PRE{} until no more
changes.  One advantage of \GCC{} implementation is that it works in sync with the
\GVN{}-\PRE{} such that when new hoisting opportunities are created by \GVN{}-\PRE{},
code-hoisting will hoist them.

@@

The algorithm for code hoisting uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Control Flow Graph (\CFG{}) and the Dominance (DOM) and Post-Dominance (PDOM)
  relations \cite{dragonbook};
\item Single Entry Single Exit (\SESE{}) \cite{sese} and Single Entry Multiple Exit
  (SEME) regions;
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given a
  unique number and the expressions that the compiler can prove identical are
  given the same number;
\item Memory \SSA{} \cite{novillo2007memory}: memory operations that the compiler
  is able to prove in dependence are linked through use-def chains.
\end{itemize}

@@

\subsection{Finding candidates to hoist}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by \GVN{}.

The current implementation of \GVN{} in \LLVM{} has some limitations when it comes to
loads and stores so we compute the \GVN{} of loads and stores separately.  Our
short-time solution to value number loads is to hash the address from where the
value is to be loaded. For stores, we value number the address as well as the
value to be stored at that address.

Another limitation of the current \GVN{} implementation in \LLVM{} is that the
instructions dependent on the loads will not get numbered correctly, and so
after hoisting all candidates we need to rerun the \GVN{} analysis in order to
discover new candidates now available after having hoisted load instructions.
This limitation should be addressed in a new implementation of the \GVN{} based on
\MemorySSA{}, that would better account for equivalent loads and their dependent
instructions.

The process of computing \GVN{} can be on-demand (as we come across an instruction)
or, precomputed (computing \GVN{} of all the instructions beforehand). Which
process to choose is determined by the scope of code-hoisting we want to
perform. In a pessimistic approach, as described in
Section~\ref{subsec:pessimistic}, we want to hoist a limited set of instructions
from the sibling branches as we iterate the DFS tree bottom-up, it is sufficient
to compute \GVN{} values on-demand. Whereas, in the optimistic approach, as
described in Section~\ref{subsec:optimistic}, we want to hoist as many
instructions as possible, and it would require \GVN{} values to be precomputed.

@@

\subsubsection{Profitability of hoisting scalars}
Since scalars are the majority of instructions which are hoisted, we pay special
attention in case of hoisting scalars too far, as that may increase register
pressure and result in spills. For example hoisting a scalar past a call, as
described in Section~\ref{cost:across-calls}.  In our current implementation we
hoist scalars past a call only when optimizing for code-side (-Os). Ideally, a
later stage of live-range splitting pass should split the live-ranges for
optimal performance, however, that is not the case with \LLVM{} as we have found
regressions when scalars are hoisted too far, as in
Section~\ref{sec:cost-models}. Another way to mitigate this problem is be to
reinstantiate (rematerialize) the computation after a call (may be as a
different optimization pass).

@@

\subsection{Code generation}
Once all the legality and profitability checks are satisfied for a set of
identical instructions, they are suitable candidates for hoisting. A copy of the
computation is inserted at the hoisting point along with any instructions which
needed to be rematerialized. Thereafter, all the computations made redundant by
the new copy are removed, and the \SSA{} form is restored by updating the
intermediate representation (IR) to reflect the changes.

After one iteration of algorithm runs through the entire function, it creates
more oppportunities for \emph{higher ranked} computations
\cite{rosen1988global}. Currently, this is a limitation of the \GVN{} analysis
pass, and so we rerun the code-hoisting algorithm until there are no more
instructions left to be hoisted.  Obviously, this is not the most optimal
approach and can be improved by ranking the computations \cite{rosen1988global},
or by improving the \GVN{} analysis to correctly number loads and dependent
instructions.

Finally after the transform is done, we verify a set of post-conditions to
establish that program invariants are maintained: e.g., consistency of
use-defs, and \SSA{} semantics.

@@

In this program the computations of tmax and tmin are identical to the
computations of tmin and tmax of sibling branch respectively. Both tmax and tmin
depends on inv which depends on a division operation which is generally more
expensive than the addition, subtraction and multiplication operations. The
total latency of computation across each branch is:
$C_{div} + 2(C_{sub} + C_{mul})$
Or, for out of order processors with two add units and two multiply units:
$C_{div} + C_{sub} + C_{mul}$

@@

In this code the two subtractions and the division operations can be executed in
parallel because there are no dependencies among them. So the total number of
cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since $C_{div}$ is
usually much greater than $C_{sub}$ \cite{x86,aarch64}.

@@

\section{Code hoisting policies}
The amount of hoisting depends on whether we collect \GVN{} of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the \GVN{} algorithm, however, that analysis is beyond
the scope of this paper.

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
In this approach, the goal is to maximize the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. After that we find the common
dominator dominating all such identical computations and perform legality
checks, as described in Section~\ref{subsec:legality}. Often times it is not
possible to hoist all the instructions to one common dominator, due to legality
constraints, e.g., intersecting side-effects, or profitability constraints, e.g.,
hoisting point too far. In those cases, this algorithmn would partition the list
of identical instructions into subsets which can be partially hoisted to their
respective common dominators. The partition algorithm is described as follows:

@@

Same property would hold for instructions I1, I2, I3 provided they are not fully
redundant, i.e., none of the basic blocks containing I1, I2 and I3 respectively
dominate the other:

@@

In the presence of fully redundant computations in the list of hoistable
candidates this equation may not hold. For example, if there is another
instruction I4 which is dominated by I1, but I4 is present in a basic block
which is farther than $DFSIn(I1, I2)$, i.e., $DFSIn(I1,I2) < DFSIn(I1,I4)$;
however, $BBDist(I1,I4) = 0 < BBDist(I1,I2) >=1$.

@@

In our current implementation we keep as many candidates in one set as possible
(greedy approach). We split the list at a point where the legality checks fail
to hoist subset of candidates which are legal to hoist and then start finding
new hoisting point for the remaining ones. So there are two limitations of the
current implementation of the partition algorithm: first, sorting by DFSIn
numbers does not give the desired order when there are fully redundant
instructions; second, it lacks cost model for partitioning in order to
preserve/improve performance, see Section~\ref{sec:future-work}.

\subsection{Hoisting bottom up with pessimistic approach}
\label{subsec:pessimistic}
In the pessimistic approach, the basic blocks are traversed in the inverse
depth-first order, computing the \GVN{} of instructions as they come by. The \GVN{} of
sibling branches are compared for equality. Once such a candidate is found, it
is hoisted in the common dominator. All the leaf nodes are visited before the
non-leaf nodes (bottom-up) because instructions are hoisted upwards.

The pessimistic algorithm is fast, results in fewer spills and also hoists fewer
instructions. We have implemented optimistic approach because it is more general
and can be tuned down to closely mimic pessimistic approach by changing a flag.

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that could
be hoisted in the program, matching the complexity of \PRE{} on \SSA{} form.  The
analysis phase is based on the Global Value Numbering (\GVN{}), the same analysis
used for \PRE{}, followed by the computation of a partition of identical
expressions to be hoisted in a same location to guarantee safety properties and
program performance, and followed by a simple code generation that adds the
identified instruction in the hoisting point and removes all the now redundant
expressions.

@@

If the new live-range is less than the old one it will be a good candidate for
hoisting. The live ranges can remain same as well if there is only one operand
on the right hand side, e.g., an assignment operation or, one of the operands is
not a register. That means hoisting upwards will decrease the live-range of its
use but increase the live-range of its definition.

@@

\subsection{In the presence of calls}
\label{cost:across-calls}
Hoisting scalars across calls is tricky because it can increase the number of
spills. During the frame lowering of calls, the argument registers, in general,
the caller saved registers are saved because they might be modified by the
callee and after the call they are restored. So before the
call, the register pressure is high because the number of available registers
are reduced by the number of caller saved registers. In that situation if a
computation is hoisted across the call, that would increase the total number of
registers required by one, thus contributing to the register pressure. However,
in the special case discussed in Section~\ref{hoist:reg-pressure}, it will be
okay to hoist because the register pressure would not decrease.

Hoisting loads/stores across calls also require precise analysis of all the
memory addresses accessed by the call. Our implementation being an
intraprocedural pass, the analysis is very conservative. In the presence of pure
calls, loads can be hoisted but stores can't. Also, if the call throws
exceptions, or if it may not return, memory references cannot be hoisted.

\subsection{Hoisting too far away}
If there are several instructions in between the hoisting point and the
instruction to be hoisted, the instruction to be hoisted crosses several
instructions while hoisting, that means we are adding one register to all the
live-ranges spanning the instructions. That could result in spills. In the
current implementation we choose to hoist if the number of instructions crossed
is below a threshold. Ideally, it should be okay to hoist all the instructions
and a later a live-range-splitting \cite{cooper1998live} pass should make the
right decision of rematerializing the instruction should it be beneficial to do
so. But the current live-range splitting pass of \LLVM{} is not making the
optimial decision and we have found spills if the threshold is exceeded. The
threshold was computed as a result of tuning the \LLVMTestSuite{}, SPEC Cpu2000,
and Cpu2006 benchmarks \cite{Henning2000}.

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
We ran \LLVMTestSuite{} (trunk:d87471f8) with the patch (trunk:86940146) and the
results are listed in Table~\ref{tab:hoist-results}. The table lists the number
of scalars, loads, stores and calls hoisted as well as removed. For each
category, the number of instructions removed is greater or equal to the number
of instructions hoisted because each hoisting is performed only when at least
one identical computation is found.

Loads are hoisted the most followed by scalars, stores and calls in decreasing
order.  This was the common trend in all our experiments. One reason why loads
are hoisted the most is the early execution of this pass (before mem2reg) in the
\LLVM{} pass pipeline. Passes like mem2reg, instcombine might actually remove
those loads so this order may change should this pass be scheduled later.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|}
      \hline
      Metric               & Number\\\hline
      Scalars hoisted      & 6791  \\\hline
      Scalars removed      & 9696  \\\hline
      Loads hoisted        & 14802 \\\hline
      Loads removed        & 20719 \\\hline
      Stores hoisted       & 15    \\\hline
      Stores removed       & 15    \\\hline
      Calls hoisted        & 8     \\\hline
      Calls removed        & 8     \\\hline
      Total Instructions hoisted & 21616 \\\hline
      Total Instructions removed & 30438 \\\hline
\end{tabular}
  \end{center}
  \caption{Code hoisting metrics on \LLVMTestSuite{}}
  \label{tab:hoist-results}
\end{table}

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|c|}
      \hline
      Metric               & Before & After              \\\hline
      Call sites deleted, not inlined             & 1988    & 1988   \\\hline
      Functions deleted (all callers found)       & 38250   & 38255  \\\hline
      Functions inlined                           & 154986  & 154985 \\\hline
      Allocas merged together                     & 212     & 212    \\\hline
      Caller-callers analyzed                     & 193042  & 193092 \\\hline
      Call sites analyzed                         & 414336  & 414381 \\\hline
      Rematerialized defs for spilling            & 18321   & 18326  \\\hline
      Rematerialized defs for splitting           & 5719    & 5842   \\\hline
      Spill slots allocated                       & 42912   & 42970  \\\hline
      Spilled live ranges                         & 61330   & 61362  \\\hline
      Spills inserted                             & 50724   & 50784  \\\hline
\end{tabular}
  \end{center}
  \caption{Static metrics before and after code-hoisting on \LLVMTestSuite{}}
  \label{tab:static-results}
\end{table}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{|l|c|}
      \hline
      Code-size metric  (.text)                   & Number   \\\hline
      Total  benchmarks                           & 497      \\\hline
      Total  gained in size                       & 39       \\\hline
      Total  decrease in size                     & 58       \\\hline
      Median decrease in size                     & 2.9\%    \\\hline
      Median increase in size                     & 2.4\%    \\\hline
    \end{tabular}
  \end{center}
  \caption{Code size metrics on \LLVMTestSuite{}}
  \label{tab:code-size}
\end{table}

While benchmarking \LLVMTestSuite{} we see both increase as well as decrease in
the codesizes of the final binaries. Since the pass runs early, it affects many
optimizations which rely on the number of instructions, lenth of the use-def
chain, and other metrics. For instance, the inliner is impacted by a decrease in
the number of instructions in the caller and callee, as its heuristics estimate
the size of functions to be inlined. Various code-size metrics are shown in
Table~\ref{tab:code-size}. All but one benchmark varied between -5.32\% and
5.43\%.  In one benchmark FreeBench/distray/distray.test, the codesize increased
by 35.38\%. In this benchmark 3 more functions got inlined (15 as compared to
12) and because of that 10 more vector instructions got generated (81 vs. 71), 3
calls got hoisted/sunk as (compared to 0), one loop got unswitched (compared to
0), 6 high latency machine instructions got hoisted out of loop, 59 (compared to
30) machine instructions got hoisted out of loop, 70 (compared to 39) machine
instructions were sunk.
